---
title: "Hands-on workshop: Instructions for R"
output: 
  html_document:
    code_folding: show
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
if(!require(magrittr)) {install.packages("magrittr"); library(magrittr);}
# rm(list = ls())
# knitr::opts_chunk$set(fig.align = 'center', fig.show='hold', out.width = '70%',
                      # warning=FALSE, message=FALSE)

knitr::opts_chunk$set(
  comment = "#>",
  # collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  warning=FALSE, message=FALSE
)
```

# Introduction 

## History

In 1991, R was created by Ross Ihaka and Robert Gentleman in the Department of Statistics at the University of Auckland. In 1993 the first announcement of R was made to the public. Ross’s and Robert’s experience developing R is documented in a 1996 paper in the Journal of Computational and Graphical Statistics:

> Ross Ihaka and Robert Gentleman. R: A language for data analysis and graphics. Journal of Computational and Graphical Statistics, 5(3):299–314, 1996

In 1995, Martin Mächler made an important contribution by convincing Ross and Robert to use the [GNU General Public License](http://www.gnu.org/licenses/gpl-2.0.html) to make R free software. This was critical because it allowed for the source code for the entire R system to be accessible to anyone who wanted to tinker with it (more on free software later).

In 1996, a public mailing list was created (the R-help and R-devel lists) and in 1997 the R Core Group was formed, containing some people associated with S and S-PLUS. Currently, the core group controls the source code for R and is solely able to check in changes to the main R source tree. Finally, in 2000 R version 1.0.0 was released to the public.

## Design of the R System

The primary R system is available from the Comprehensive R Archive Network, also known as CRAN. CRAN also hosts many add-on packages that can be used to extend the functionality of R.

The R system is divided into 2 conceptual parts:

1. The “base” R system that you download from CRAN: Linux Windows Mac Source Code
1. Everything else.

R functionality is divided into a number of packages.

- The “base” R system contains, among other things, the `base` package which is required to run R and contains the most fundamental functions.
- The other packages contained in the “base” system include `utils`, `stats`, `datasets`, `graphics`, `grDevices`, `grid`, `methods`, `tools`, `parallel`, `compiler`, `splines`, `tcltk`, `stats4`.
- There are also “Recommended” packages: `boot`, `class`, `cluster`, `codetools`, `foreign`, `KernSmooth`, `lattice`, `mgcv`, `nlme`, `rpart`, `survival`, `MASS`, `spatial`, `nnet`, `Matrix`.

When you download a fresh installation of R from CRAN, you get all of the above, which represents a substantial amount of functionality. However, there are many other packages available:

- There are over 4000 packages on CRAN that have been developed by users and programmers around the world. 
    - <span style="color:red">**Remark for Hands-on session :** We will manually install and use `bnlearn` package from CRAN</span>
- There are also many packages associated with the [Bioconductor project](http://bioconductor.org/). 
    - <span style="color:red">**Remark for Hands-on session** : We will manually install and use `Rgraphviz` package from [Bioconductor project](http://bioconductor.org/)</span>

----

----

# Obtaining R

To download R, go to CRAN, the **c**omprehensive **R** **a**rchive **n**etwork. CRAN is composed of a set of mirror servers distributed around the world and is used to distribute R and R packages. Don't try and pick a mirror that's close to you: instead use the cloud mirror, <https://cloud.r-project.org>, which automatically figures it out for you.

A new major version of R comes out once a year, and there are 2-3 minor releases each year. It's a good idea to update regularly. Upgrading can be a bit of a hassle, especially for major versions, which require you to reinstall all your packages, but putting it off only makes it worse.

<span style="color:red">**For our hands-on sessions in Konstanz, download R version 3.4.3 is recommended.**</span>

## Install R for Windows

1. Download latest R for Windows : go to https://cloud.r-project.org/bin/windows/base/release.htm
    - Download should automatically start
    - Otherwise go to : http://cran.r-project.org/bin/windows/base/ and manually [download R version 3.4.3](https://cran.r-project.org/bin/windows/base/R-3.4.3-win.exe)
1. Install R : Leave all default settings in the installation options

Note: c.f. <button class="btn btn-primary" data-toggle="collapse" data-target="#YoutubeInstallRWin"> Video : "How to Install R"</button>
<div id="YoutubeInstallRWin" class="collapse">  
<div class="boxed">

<iframe width="630" height="472"
src="https://www.youtube.com/embed/MSmWl-sbn64">
</iframe>

</div>
</div>


## Install R for Mac OS

If you are a user of **homebrew**, we do NOT recommend you install R via `homebrew`.

1. Download latest R for Mac OS :
    - Download the latest R 
        - Go to [this page](https://cran.r-project.org/bin/macosx/) https://cran.r-project.org/bin/macosx/ and [download R-3.4.3.pkg](https://cloud.r-project.org/bin/macosx/R-3.4.3.pkg) (latest R for Konstanz Summer School)
    - Download [XQuartz](https://www.xquartz.org/)
        - go to [this page](https://www.xquartz.org/) https://www.xquartz.org/ and [download XQuartz-2.7.11.dmg](https://dl.bintray.com/xquartz/downloads/XQuartz-2.7.11.dmg)
        - **Note** :the use of X11 (including `tcltk`) requires `XQuartz` to be installed since it is no longer part of OS X. Always re-install XQuartz when upgrading your OS X to a new major version.
1. Install R : Leave all default settings in the installation options.
1. Install XQuartz : Leave all default settings in the installation options.

Note: c.f. <button class="btn btn-primary" data-toggle="collapse" data-target="#YoutubeInstallRMac"> Video “Installing XQuartz and R for Mac” </button> 
<div id="YoutubeInstallRMac" class="collapse">  
<div class="boxed">

<iframe width="630" height="472"
src="https://www.youtube.com/embed/6-I0nd5ojXM">
</iframe>

</div>
</div>


```{r, eval=FALSE, include=FALSE}
c(420,315)*1.5
```


----

----

# Getting Started with R

TODO

----
<!-- ### R Session { - } -->

<!-- After R is started, there is a console awaiting for input. At the prompt `>`, you can enter numbers and perform calculations. -->

<!-- ```{r} -->
<!-- 1 + 2  -->
<!-- ``` -->

<!-- ### Variable Assignment { - }  -->

<!-- We assign values to variables with the assignment operator `=`. Just typing the variable by itself at the prompt will print out the value. We should note that another form of assignment operator `<-` is also in use. -->

<!-- ```{r} -->
<!-- x = 1 -->
<!-- x  -->
<!-- ``` -->

<!-- ### Functions { - } -->

<!-- R functions are invoked by its name, then followed by the parenthesis, and zero or more arguments. The following apply the function `c()` to combine three numeric values into a vector. -->

<!-- ```{r} -->
<!-- c(1, 2, 3)  -->
<!-- ``` -->

<!-- ### Comments { - } -->

<!-- All text after the pound sign "#" within the same line is considered a comment. -->

<!-- ```{r} -->
<!-- 1 + 1      # this is a comment -->
<!-- ``` -->

<!-- ### Extension Package { - } -->

<!-- Sometimes we need additional functionality beyond those offered by the core R library. In order to install an extension package, you should invoke the `install.packages()` function at the prompt and follow the instruction. We will need `bnlearn` package. -->

<!-- ```{r eval = FALSE} -->
<!-- install.packages("bnlearn") # install bnlearn package -->
<!-- ``` -->

<!-- ### Getting Help { - }  -->

<!-- R provides extensive documentation. For example, entering `?c` or `help(c)` at the prompt gives documentation of the function `c` in R. Please give it a try. -->

<!-- ```{r} -->
<!-- help(c) -->
<!-- ``` -->

<!-- ---- -->

<!-- To run a typed command, just hit enter -->

<!-- The place in the RGui where you input commands is indicated in -->
<!-- the screenshot below:  -->



<!-- ---- -->

<!-- what is console ? -->
<!-- it begins with a `>` -->

<!-- # packages -->

<!-- # Basics -->

<!-- ## Object -->

<!-- vector -->
<!-- matrix -->
<!-- data.frame -->

<!-- plot() -->
<!-- ls() -->



<!-- df <- data.frame(x = c(1, 2, 3), y = c(0, 10, NA), z= c("a",NA,"b")) -->
<!-- df -->
<!-- na.omit(df) -->
<!-- na.rm(df) -->
<!-- na.rm(df, c("x","y")) -->
<!-- na.rm(df, "z") -->

<!-- library(questionr) -->

<!-- x <- list(abc = 1) -->
